A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-23 03-semantics.tex
Seção 3.3, terminação (início)

\subsection{Determinism, Termination, and Memory Bounds}

@@

Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
description either no rule is applicable or exactly one of them can be
applied.  This means that the resulting relation~$\trans$ is in fact a
partial function and that, consequently, the same number of applications
of~$\trans$ will always yield the same result.

The next two lemmas establish the determinism of a single application
of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
inspection of rules~\R{push} and~\R{pop}.  The proof of
Lemma~\ref{lem.x.det-out} proceeds by induction on the structure of the
derivation trees produced by applications of the rules for~$\nst$.  Both
lemmas are used in the proof of Theorem~\ref{thm.x.det}, the main result of
this section, which establishes that any given number of applications
of~$\trans$ starting from the same input description will always lead to the
same output description.

@@

\subsubsection*{Termination}

- there is always a possible transition until n=0

every cannot restart itself
    - break disallowed
    - emit ignored

@@

>>>>

\subsection{Properties}

@@

Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
description either no rule is applicable or exactly one of them can be
applied.  This means that the resulting relation~$\trans$ is in fact a
partial function.

The next two lemmas establish the determinism of a single application
of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
inspection of rules~\R{push} and~\R{pop}.  The proof of
Lemma~\ref{lem.x.det-out}, however, requires an induction on the structure
of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
used in the proof of Theorem~\ref{thm.x.det}, the main result of this
section.  Theorem~\ref{thm.x.det} establishes that any given number of
applications of~$\trans$ starting from the same input description will
always lead to the same output description.

@@

\subsubsection{Termination}

In this section, we prove that a sufficiently long sequence of applications
of~$\trans$, and consequently~$\out$ and~$\nst$, will eventually lead to an
irreducible description, viz., one that cannot be modified by further
transitions.

\begin{definition}
  \label{def.x.Hnst}
  %%
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
  true.
\end{definition}

Nested-irreducible descriptions serve as normal forms for~$\nst$
transitions: they embody the result of an exhaustive number of~$\nst$
applications.  We will write~$\delta_\Hnst$ to indicate that
description~$\delta$ is nested-irreducible.

To characterize irreducible descriptions in general, we will need to define
the notions of potency of a program and the rank of a description.

\begin{definition}
  \label{def.x.pot}
  %%
  The potency of a program~$p$ in reaction to event~$e$,
  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
  that can be executed during a reaction of~$p$ to~$e$.

  More formally,
  \[
    \pot(p,e)=\pot'(\bcast(p,e))\,,
  \]
  where~$\pot'$ is an auxiliary function that counts the number of
  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
  broadcast of event~$e$ to~$p$.

  The auxiliary function~$\pot'$ is defined by the following clauses:
  \begin{enumerate}[(a)]
  \item\label{def.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$;
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
    =\max\{\pot'(p_1),\pot'(p_2)\}$;
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$;
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$;
    %%
  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$;
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$'';
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$,
    \[
      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
    \]
    %%
  \item\label{def.pot.last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
    \]
    %%
  \item Otherwise, if none of~\eqref{def.pot.first}--\eqref{def.pot.last}
    applies, $\pot(\_)=0$.
  \end{enumerate}
\end{definition}

\begin{definition}
  \label{def.x.rank}
  %%
  The \emph{rank} of a description~$\delta=\<p,n,e>$,
  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
  that
  \begin{alignat*}{2}
    i&=\pot(p,e) &\quad\text{and}\quad
    j&=
       \begin{cases}
         n  &\text{if~$e=\nil$}\\
         n+1&\text{otherwise\,.}
       \end{cases}
  \end{alignat*}
\end{definition}

\begin{definition}
  \label{def.x.H}
  %%
  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
  some~$i\ge0$.
\end{definition}

An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
transitions~$\trans$ in general.  Such descriptions cannot be advanced
by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
nor~$\outpop$, as the second coordinate of its rank is~0, which
implies~$e=\nil$ and~$n=0$.

TODO: Lemas e teorema da terminação.  Sketch da prova.  Mencionar que as
restrições sintáticas são fundamentais.

% - there is always a possible transition until n=0
%
% every cannot restart itself
%     - break disallowed
%     - emit ignored

@@

TODO: Acho que sai direto do~$\pot(p)$.
